---
title: "Case_Study"
author: "Yiqian"
date: "2025-06-13"
output: html_document
---

# Sources

```{r}
library(phyloseq)
library(vegan)
library(ggplot2)
library(ggpubr)
library(car)
library(quantreg)
library(dplyr)
library(ggsignif)
library(hillR)
# devtools::install_github("bioscinema/WasserSpectrum")
library(WasserSpectrum)
```


# CVD

```{r}
load("./Real Data/cardiovascular_wgs.RData")
meta <- as(sample_data(ps),"data.frame")
meta$SampleID <- rownames(meta)

meta <- meta %>%
  filter(grepl("^MMC", Status) | grepl("^HC", Status) | grepl("^IHD", Status)) %>%
  mutate(Status = ifelse(grepl("^MMC", Status), "MMC",
                         ifelse(grepl("^HC", Status), "HC", 
                                ifelse(grepl("^IHD", Status), "IHD", NA))))

meta <- meta %>%
  mutate(HbA1c.... = ifelse(HbA1c.... == "NA", NA, HbA1c....)) %>% 
  filter(!is.na(HbA1c....)) %>%
  mutate(HbA1c = as.numeric(HbA1c....))

meta <- meta %>%
  filter(Gender %in% c("Female", "Male"))

ps_filt <- prune_samples(meta$SampleID, ps)
sample_data(ps_filt) <- sample_data(meta)
ps_filt <- prune_taxa(taxa_sums(ps_filt) > 0, ps_filt)

ps_filt@sam_data$Age = ps_filt@sam_data$Age..years.
```


## Shannon

```{r}
shannon_vec <- vegan::diversity(t(otu_table(ps_filt)), index = "shannon")

# Get levels of Status
status_levels <- unique(sample_data(ps_filt)$Status)

# Store results in a named list
shannon_by_status <- setNames(vector("list", length(status_levels)), status_levels)

# Loop through each Status group
for (s in status_levels) {
  ps_sub <- subset_samples(ps_filt, Status == s)
  shannon_by_status[[s]] <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
}

```


### Part 1: Three Group Comparison (traditional analysis)

```{r, warning=FALSE}
meta_df <- data.frame(sample_data(ps_filt))
meta_df$SampleID <- rownames(meta_df)
meta_df$Shannon <- shannon_vec[meta_df$SampleID]
meta_df$Depth <- colSums(otu_table(ps_filt))

kruskal_res <- kruskal.test(Shannon ~ Status, data = meta_df)
p_val <- format.pval(kruskal_res$p.value, digits = 4, eps = .0001)

ggplot(meta_df, aes(x = Status, y = Shannon, fill = Status)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.7) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Alpha Diversity by Group (Shannon Index)",
    y = "Shannon Index",
    x = "Status"
  ) +
  annotate("text", x = 1.5, y = max(meta_df$Shannon) * 1.05,
           label = paste0("Kruskal-Wallis p = ", p_val), size = 4.5) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )

frechet_wasserstein_test(meta_df, 
                         "Shannon", 
                         "Status", 
                         nperm = 1000, 
                         plot = TRUE,
                         seed = 123)

```

### Part 2: Wasserstein Spectrum of HbA1c
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
print(status_levels)
variable <- "HbA1c"
shannon_by_status <- list()
meta_by_status <- list()
status_levels = c("MMC", "HC")
for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, ".eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})



```

#### Negative Control

```{r}
wasserstein_spectrum_nca <- function(df, 
                                     feature_col, 
                                     outcome_col, 
                                     confounder_cols = NULL,
                                     basis_df = 6,
                                     t_grid = seq(0.01, 0.99, length.out = 100),
                                     n_perm = 99,
                                     seed = 123) {
  
  if (!is.null(seed)) set.seed(seed)
  
  # Extract data
  y <- df[[feature_col]]
  x <- df[[outcome_col]]
  if (is.factor(x)) {
    if (nlevels(x) != 2) stop("outcome_col must be binary or numeric.")
    x <- as.numeric(x) - 1
  } else if (!is.numeric(x)) {
    stop("outcome_col must be numeric or a binary factor.")
  }
  conf_df <- if (!is.null(confounder_cols)) df[, confounder_cols, drop = FALSE] else NULL
  
  n <- nrow(df)
  m <- length(t_grid)
  
  # Build quantile thresholds
  quantile_thresholds <- quantile(y, probs = t_grid, type = 8)
  
  # Construct binary matrix: n x m, then flatten
  y_bin_mat <- outer(y, quantile_thresholds, FUN = ">") * 1  # n x m
  y_vec <- as.vector(y_bin_mat)
  
  # Full model design matrix (with x)
  model_data <- data.frame(x = x)
  if (!is.null(conf_df)) model_data <- cbind(model_data, conf_df)
  X0 <- model.matrix(~ ., data = model_data)  # n x p
  p <- ncol(X0)
  
  # Reduced model design matrix (without x)
  model_data_red <- if (!is.null(conf_df)) conf_df else data.frame(intercept = rep(1, n))
  X0_red <- model.matrix(~ ., data = model_data_red)  # n x p_red
  
  # B-spline basis on t_grid
  Phi <- splines::bs(t_grid, df = basis_df, intercept = TRUE)  # m x K
  K <- ncol(Phi)
  
  # Full model: fit and extract beta1(t)
  X_big <- kronecker(Phi, X0)          # (n*m) x (p*K)
  fit_full <- lm(y_vec ~ X_big - 1)
  coef_hat <- coef(fit_full)
  
  idx_x <- which(colnames(X0) == "x")
  if (length(idx_x) == 0) stop("Couldn't find predictor `x` in design matrix.")
  sel_idx <- idx_x + (0:(K-1)) * p
  
  beta1_grid <- numeric(m)
  for (j in 1:m) {
    beta1_grid[j] <- sum(coef_hat[sel_idx] * Phi[j, ])
  }
  
  # Reduced model: get fitted values and residuals
  X_big_red <- kronecker(Phi, X0_red)  # (n*m) x (p_red*K)
  fit_red <- lm(y_vec ~ X_big_red - 1)
  fitted_red <- fitted(fit_red)        # length n*m
  resid_red  <- residuals(fit_red)     # length n*m
  
  # Reshape residuals to n x m matrix for blocked permutation
  resid_mat <- matrix(resid_red, nrow = n, ncol = m)  # n x m
  
  # Permutation: store beta1 for each permutation
  perm_beta1 <- matrix(NA, nrow = n_perm, ncol = m)
  
  for (b in 1:n_perm) {
    # Blocked permutation: permute rows (samples), keep columns (quantiles) intact
    perm_idx <- sample(n)
    resid_perm <- resid_mat[perm_idx, ]       # n x m, rows permuted
    
    # Reconstruct permuted y_vec
    y_vec_perm <- fitted_red + as.vector(resid_perm)  # n*m
    
    # Fit full model on permuted response
    fit_perm <- lm(y_vec_perm ~ X_big - 1)
    coef_perm <- coef(fit_perm)
    
    for (j in 1:m) {
      perm_beta1[b, j] <- sum(coef_perm[sel_idx] * Phi[j, ])
    }
  }
  
  return(list(
    quantiles   = t_grid,
    beta1       = beta1_grid,
    perm_beta1  = perm_beta1   # n_perm x m matrix
  ))
}


plot_nca <- function(result,
                     alpha = 0.1,
                     n_ticks = 10,
                     title = "Negative Control Analysis (Blocked Freedman-Lane Permutation)",
                     xlab = expression(tau),
                     ylab = "Coefficient",
                     box_color = "grey70",
                     point_color = "red",
                     point_shape = 4,
                     point_size = 1.5,
                     bound_color = "blue",
                     bound_shape = 1,
                     bound_size = 1.5) {
  
  t_grid     <- result$quantiles
  beta1      <- result$beta1
  perm_beta1 <- result$perm_beta1  # n_perm x m
  
  m <- length(t_grid)
  
  # Compute permutation quantile bounds at each quantile level
  perm_lower <- apply(perm_beta1, 2, quantile, probs = alpha / 2)
  perm_upper <- apply(perm_beta1, 2, quantile, probs = 1 - alpha / 2)
  
  # Build long data frame for boxplot (all quantile levels)
  perm_long <- data.frame(
    tau  = rep(t_grid, each = nrow(perm_beta1)),
    coef = as.vector(perm_beta1)
  )
  perm_long$tau <- factor(perm_long$tau)
  
  # Observed beta and bounds at all quantile levels
  obs_df <- data.frame(
    tau   = factor(t_grid),
    coef  = beta1,
    lower = perm_lower,
    upper = perm_upper
  )
  
  # Select n_ticks evenly spaced levels for x-axis labels
  tick_idx    <- round(seq(1, m, length.out = n_ticks))
  tick_labels <- levels(perm_long$tau)
  show_labels <- ifelse(
    seq_along(tick_labels) %in% tick_idx,
    round(t_grid, 2),
    ""
  )
  
  p <- ggplot(perm_long, aes(x = tau, y = coef)) +
    geom_boxplot(
      fill = box_color, color = "grey40",
      outlier.shape = NA,
      width = 0.6
    ) +
    geom_point(
      data  = obs_df,
      aes(x = tau, y = coef),
      color = point_color,
      shape = point_shape,
      size  = point_size
    ) +
    geom_point(
      data  = obs_df,
      aes(x = tau, y = lower),
      color = bound_color,
      shape = bound_shape,
      size  = bound_size
    ) +
    geom_point(
      data  = obs_df,
      aes(x = tau, y = upper),
      color = bound_color,
      shape = bound_shape,
      size  = bound_size
    ) +
    scale_x_discrete(labels = show_labels) +
    labs(
      title = title,
      x     = xlab,
      y     = ylab
    ) +
    theme_classic(base_size = 13) +
    theme(
      plot.title   = element_text(hjust = 0.5, face = "bold", size = 13),
      axis.title.x = element_text(size = 11, margin = margin(t = 6)),
      axis.title.y = element_text(size = 11, margin = margin(r = 6)),
      axis.text    = element_text(color = "black")
    )
  
  return(p)
}


library(tidyverse)

B <- 1000
set.seed(1)

t_grid_show <- seq(0.01, 0.99, length.out = 100)

models <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "HbA1c",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = t_grid_show,
  n_perm = B,
  seed = 123
)

# keep old-style objects if you still want them later
df_tau_coef <- data.frame(
  tau  = models$quantiles,
  coef = models$beta1
)

perm_mat <- t(models$perm_beta1)   # now m x B, so rows match tau
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))
df_tau_coef <- cbind(df_tau_coef, perm_mat)

df_perm_long <- df_tau_coef %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

p <- plot_nca(
  result = models,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis",
  xlab = expression(tau),
  ylab = "Coefficient",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "HbA1c_NegativeControl.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```


```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

# Define grid and number of permutations
taus <- seq(0.01, 0.99, length.out = 100)
B <- 1000
set.seed(1)

# Run the new NCA model with confounder adjustment
obs_model <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "HbA1c",
  confounder_cols = c("Depth", "Age", "Gender"),
  basis_df = 6,
  t_grid = taus,
  n_perm = B,
  seed = 123
)

# Keep an observed coefficient data frame
df_tau_coef <- data.frame(
  tau  = obs_model$quantiles,
  coef = obs_model$beta1
)

# Convert permutation matrix to old-style format:
# perm_beta1 is n_perm x n_tau, so transpose to n_tau x n_perm
perm_mat <- t(obs_model$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

# Bind observed and permuted coefficients
df_final <- cbind(df_tau_coef, perm_mat)

# Long format for any downstream use
df_perm_long <- df_final %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# Plot using the new plotting function
p <- plot_nca(
  result = obs_model,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis (Blocked Freedman-Lane Permutation)",
  xlab = expression(tau),
  ylab = "Coefficient (HbA1c)",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p)

# Save figure
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "HbA1c_NegativeControl_Confounder.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)

```


### Part 3: Wasserstein Spectrum of Status

```{r, warning=FALSE}
Status.ws <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws$plot



### adjust sequencing depth

Status.ws2 <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws2$plot

### adjust sequencing depth and other covariates

Status.ws3 <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = c("Depth","Age","HbA1c","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws3$plot


```

#### Negative Control

```{r}
library(tidyverse)
B <- 1000
set.seed(1)

t_grid_show <- seq(0.01, 0.99, length.out = 100)

# observed fit
fit_obs <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = t_grid_show,
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

df_tau_coef <- data.frame(
  tau  = fit_obs$quantiles,
  coef = fit_obs$beta[1,]
)

# permutation null
status_levels <- levels(as.factor(meta_df$Status))
perm_mat <- replicate(B, {
  dfp <- meta_df
  dfp$Status <- factor(sample(as.character(dfp$Status)),
                       levels = status_levels)

  fit_p <- wasserstein_spectrum_multiclass(
    df = dfp,
    feature_col = "Shannon",
    outcome_col = "Status",
    reference_level = "IHD",
    confounder_cols = NULL,
    basis_df = 6,
    t_grid = t_grid_show,
    alpha = 0.05,
    plot = FALSE,
    seed = NULL   # or set a per-rep seed if needed
  )

  fit_p$beta[1,]
})

colnames(perm_mat) <- paste0("Rep_", seq_len(B))
df_tau_coef <- cbind(df_tau_coef, perm_mat)

df_perm_long <- df_tau_coef %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# 1) Define ordered tau levels (numeric) and choose 10 to label
tau_levels <- sort(unique(df_perm_long$tau))
idx_10 <- unique(round(seq(1, length(tau_levels), length.out = 10)))
lab_levels <- tau_levels[idx_10]

# 2) Plot: all boxplots, but only 10 x-axis labels
p1 <- ggplot() +
  geom_boxplot(
    data = df_perm_long,
    aes(x = factor(tau, levels = tau_levels), y = coef_perm),
    fill = "grey80",
    color = "grey40",
    outlier.shape = NA,
    width = 0.6
  ) +
  geom_point(
    data = df_tau_coef,
    aes(x = factor(tau, levels = tau_levels), y = coef),
    color = "red",
    shape = 4,
    size = 3,
    stroke = 1.2
  ) +
  scale_x_discrete(
    breaks = as.character(lab_levels),
    labels = function(x) sprintf("%.2f", as.numeric(x))
  ) +
  labs(
    x = expression(tau),
    y = "Coefficient",
    title = "Negative Control Analysis"
  ) +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p1)

ggplot2::ggsave(
    filename = file.path("figures", "Status_NegativeControl.png"),
    plot = p1,
    width = 8, height = 6, units = "in",
    dpi = 300,
    bg = "white"
  )
```


```{r}
library(quantreg)
library(tidyverse)

taus <- seq(0.01, 0.99, length.out = 100)
B <- 1000
set.seed(1)

# Ensure IHD is the reference
meta_df <- meta_df %>%
  mutate(Status = relevel(factor(Status), ref = "IHD"))

form_red  <- Shannon ~ Age + Gender + Depth
form_full <- Shannon ~ Status + Age + Gender + Depth

# Helper: extract coefficient for a specific contrast (e.g., "HC" or "MMC")
get_contrast_coef <- function(fit, contrast_level) {
  nm <- paste0("Status", contrast_level)
  b <- coef(fit)
  if (!nm %in% names(b)) {
    stop("Coefficient ", nm, " not found. Available Status terms: ",
         paste(grep("^Status", names(b), value = TRUE), collapse = ", "))
  }
  unname(b[nm])
}

run_freedman_lane_qr <- function(contrast_level, taus, B, meta_df, form_red, form_full) {

  # Observed coefficients
  obs_coef <- sapply(taus, function(tau) {
    fit <- rq(form_full, data = meta_df, tau = tau)
    get_contrast_coef(fit, contrast_level)
  })
  df_tau_coef <- data.frame(tau = taus, coef = obs_coef)

  # Permutation matrix: n_tau x B
  perm_mat <- sapply(seq_along(taus), function(k) {
    tau <- taus[k]

    fit0  <- rq(form_red, data = meta_df, tau = tau)
    yhat0 <- fitted(fit0)
    e0    <- resid(fit0)

    replicate(B, {
      y_star <- yhat0 + sample(e0)
      df_star <- meta_df
      df_star$Shannon <- y_star

      fit1 <- rq(form_full, data = df_star, tau = tau)
      get_contrast_coef(fit1, contrast_level)
    })
  })

  perm_mat <- t(perm_mat)
  colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

  df_final <- cbind(df_tau_coef, perm_mat)

  df_perm_long <- df_final %>%
    select(tau, starts_with("Rep_")) %>%
    pivot_longer(
      cols = starts_with("Rep_"),
      names_to = "Rep",
      values_to = "coef_perm"
    )

  # Only show 10 tau labels
  tau_levels <- sort(unique(df_perm_long$tau))
  idx_10 <- unique(round(seq(1, length(tau_levels), length.out = 10)))
  lab_levels <- tau_levels[idx_10]

  p <- ggplot() +
    geom_boxplot(
      data = df_perm_long,
      aes(x = factor(tau, levels = tau_levels), y = coef_perm),
      fill = "grey80",
      color = "grey40",
      outlier.shape = NA,
      width = 0.6
    ) +
    geom_point(
      data = df_final,
      aes(x = factor(tau, levels = tau_levels), y = coef),
      color = "red",
      shape = 4,
      size = 3,
      stroke = 1.2
    ) +
    scale_x_discrete(
      breaks = as.character(lab_levels),
      labels = function(x) sprintf("%.2f", as.numeric(x))
    ) +
    labs(
      x = expression(tau),
      y = "Coefficient",
      title = paste0("Negative Control Analysis (Freedman–Lane): ", contrast_level, " vs IHD")
    ) +
    theme_classic(base_size = 13) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  list(df_final = df_final, df_perm_long = df_perm_long, plot = p)
}

# Run for HC and MMC
res_HC  <- run_freedman_lane_qr("HC",  taus, B, meta_df, form_red, form_full)
res_MMC <- run_freedman_lane_qr("MMC", taus, B, meta_df, form_red, form_full)

p1 <- res_HC$plot
p2 <- res_MMC$plot

print(p1)
print(p2)

ggplot2::ggsave(
    filename = file.path("figures", "Status_NegativeControl_ConfounderHC.png"),
    plot = p1,
    width = 8, height = 6, units = "in",
    dpi = 300,
    bg = "white"
  )
ggplot2::ggsave(
    filename = file.path("figures", "Status_NegativeControl_ConfounderMMC.png"),
    plot = p2,
    width = 8, height = 6, units = "in",
    dpi = 300,
    bg = "white"
  )
```


### Part 4: Wasserstein Spectrum of BMI
```{r, warning=FALSE}
shannon_by_status <- list()
meta_by_status <- list()
variable <- "BMI"
status_levels = c("MMC", "HC")
for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$BMI..kg.m.. <- as.numeric(meta$BMI..kg.m..)
  meta <- meta[!is.na(meta$BMI..kg.m..), ]
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, ".eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})




```

#### Negative Control

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

B <- 1000
set.seed(1)

t_grid_show <- seq(0.01, 0.99, length.out = 100)

# Make sure BMI is numeric and remove missing values
meta_df$BMI..kg.m.. <- as.numeric(meta_df$BMI..kg.m..)
meta_df <- meta_df %>%
  dplyr::filter(!is.na(BMI..kg.m..))

# Run the new NCA model
models <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "BMI..kg.m..",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = t_grid_show,
  n_perm = B,
  seed = 123
)

# Keep old-style observed coefficient data frame
df_tau_coef <- data.frame(
  tau  = models$quantiles,
  coef = models$beta1
)

# Convert permutation matrix to old-style format
# models$perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(models$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

df_tau_coef <- cbind(df_tau_coef, perm_mat)

# Long format, same as your previous workflow
df_perm_long <- df_tau_coef %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# Plot using the new plotting function
p <- plot_nca(
  result = models,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis",
  xlab = expression(tau),
  ylab = "Coefficient",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "BMI_NegativeControl.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```


```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

# Define grid and number of permutations
taus <- seq(0.01, 0.99, length.out = 100)
B <- 1000
set.seed(1)

# Make sure BMI is numeric and remove missing values
meta_df$BMI..kg.m.. <- as.numeric(meta_df$BMI..kg.m..)

meta_df <- meta_df %>%
  dplyr::filter(!is.na(BMI..kg.m..))

# Run the new NCA model with confounder adjustment
obs_model <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "BMI..kg.m..",
  confounder_cols = c("Depth", "Age", "Gender"),
  basis_df = 6,
  t_grid = taus,
  n_perm = B,
  seed = 123
)

# Observed coefficients
df_tau_coef <- data.frame(
  tau  = obs_model$quantiles,
  coef = obs_model$beta1
)

# Convert permutation matrix to old-style format
# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(obs_model$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

# Bind observed coefficients with permuted coefficients
df_final <- cbind(df_tau_coef, perm_mat)

# Pivot to long format for any downstream use
df_perm_long <- df_final %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# Plot
p <- plot_nca(
  result = obs_model,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis (Blocked Freedman-Lane Permutation)",
  xlab = expression(tau),
  ylab = "Coefficient (BMI)",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p)

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "BMI_NegativeControl_confounder.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```

### Part 5: Wasserstein Spectrum of Waist.circumference
```{r, warning=FALSE}
shannon_by_status <- list()
meta_by_status <- list()

for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Waist.circumference..cm. <- as.numeric(meta$Waist.circumference..cm.)
  meta <- meta[!is.na(meta$Waist.circumference..cm.), ]
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <- lapply(meta_by_status, function(df_sub) {
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Waist.circumference..cm.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Waist.circumference..cm.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Waist.circumference..cm.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
})



```

### Part 6: Wasserstein Spectrum of Fasting.plasma.CRP
```{r, warning=FALSE}
shannon_by_status <- list()
meta_by_status <- list()

for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.CRP..mg.L. <- as.numeric(meta$Fasting.plasma.CRP..mg.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.CRP..mg.L.), ]
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <- lapply(meta_by_status, function(df_sub) {
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.CRP..mg.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.CRP..mg.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.CRP..mg.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
})

```


### Part 7: Wasserstein Spectrum of pro.ANP
```{r, warning=FALSE}
shannon_by_status <- list()
meta_by_status <- list()
status_levels = c("HC", "IHD")
variable <- "ANP"
for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.pro.ANP..pmol.L. <- as.numeric(meta$Fasting.plasma.pro.ANP..pmol.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.pro.ANP..pmol.L.), ]
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, ".eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})


```

#### Negative Control

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

B <- 1000
set.seed(1)

t_grid_show <- seq(0.01, 0.99, length.out = 100)

meta_df$Fasting.plasma.pro.ANP..pmol.L. <- as.numeric(meta_df$Fasting.plasma.pro.ANP..pmol.L.)

meta_df <- meta_df %>%
  dplyr::filter(!is.na(Fasting.plasma.pro.ANP..pmol.L.))

models <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = t_grid_show,
  n_perm = B,
  seed = 123
)

df_tau_coef <- data.frame(
  tau  = models$quantiles,
  coef = models$beta1
)

# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(models$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

df_tau_coef <- cbind(df_tau_coef, perm_mat)

df_perm_long <- df_tau_coef %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

p <- plot_nca(
  result = models,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis",
  xlab = expression(tau),
  ylab = "Coefficient",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "proANP_NegativeControl.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

# Define grid and number of permutations
taus <- seq(0.01, 0.99, length.out = 100)
B <- 1000
set.seed(1)

# Make sure outcome is numeric and remove missing values
meta_df$Fasting.plasma.pro.ANP..pmol.L. <- as.numeric(
  meta_df$Fasting.plasma.pro.ANP..pmol.L.
)

meta_df <- meta_df %>%
  dplyr::filter(!is.na(Fasting.plasma.pro.ANP..pmol.L.))

# Run the new NCA model with confounder adjustment
obs_model <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.",
  confounder_cols = c("Depth", "Age", "Gender"),
  basis_df = 6,
  t_grid = taus,
  n_perm = B,
  seed = 123
)

# Observed coefficients
df_tau_coef <- data.frame(
  tau  = obs_model$quantiles,
  coef = obs_model$beta1
)

# Convert permutation matrix to old-style format
# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(obs_model$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

# Bind observed coefficients with permuted coefficients
df_final <- cbind(df_tau_coef, perm_mat)

# Pivot to long format for any downstream use
df_perm_long <- df_final %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# Plot
p <- plot_nca(
  result = obs_model,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis (Blocked Freedman-Lane Permutation)",
  xlab = expression(tau),
  ylab = "Coefficient (proANP)",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p)

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "proANP_NegativeControl_confounder.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```


### Part 8: Wasserstein Spectrum of SBP
```{r, warning=FALSE}
shannon_by_status <- list()
meta_by_status <- list()
status_levels = c("HC", "IHD")
variable <- "SBP"
for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Systolic.blood.pressure..mmHg. <- as.numeric(meta$Systolic.blood.pressure..mmHg.)
  meta <- meta[!is.na(meta$Systolic.blood.pressure..mmHg.), ]
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, ".eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})



```


#### Negative Control

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

B <- 1000
set.seed(1)

t_grid_show <- seq(0.01, 0.99, length.out = 100)

meta_df$Systolic.blood.pressure..mmHg. <- as.numeric(meta_df$Systolic.blood.pressure..mmHg.)

meta_df <- meta_df %>%
  dplyr::filter(!is.na(Systolic.blood.pressure..mmHg.))

models <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Systolic.blood.pressure..mmHg.",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = t_grid_show,
  n_perm = B,
  seed = 123
)

df_tau_coef <- data.frame(
  tau  = models$quantiles,
  coef = models$beta1
)

# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(models$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

df_tau_coef <- cbind(df_tau_coef, perm_mat)

df_perm_long <- df_tau_coef %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

p <- plot_nca(
  result = models,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis",
  xlab = expression(tau),
  ylab = "Coefficient",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "SBP_NegativeControl.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

# Define grid and number of permutations
taus <- seq(0.01, 0.99, length.out = 100)
B <- 1000
set.seed(1)

# Make sure outcome is numeric and remove missing values
meta_df$Systolic.blood.pressure..mmHg. <- as.numeric(
  meta_df$Systolic.blood.pressure..mmHg.
)

meta_df <- meta_df %>%
  dplyr::filter(!is.na(Systolic.blood.pressure..mmHg.))

# Run the new NCA model with confounder adjustment
obs_model <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Systolic.blood.pressure..mmHg.",
  confounder_cols = c("Depth", "Age", "Gender"),
  basis_df = 6,
  t_grid = taus,
  n_perm = B,
  seed = 123
)

# Observed coefficients
df_tau_coef <- data.frame(
  tau  = obs_model$quantiles,
  coef = obs_model$beta1
)

# Convert permutation matrix to old-style format
# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(obs_model$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

# Bind observed coefficients with permuted coefficients
df_final <- cbind(df_tau_coef, perm_mat)

# Pivot to long format for any downstream use
df_perm_long <- df_final %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# Plot
p <- plot_nca(
  result = obs_model,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis (Blocked Freedman-Lane Permutation)",
  xlab = expression(tau),
  ylab = "Coefficient (SBP)",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p)

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "SBP_NegativeControl_confounder.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```

### Part 9: Wasserstein Spectrum of DBP
```{r, warning=FALSE}
shannon_by_status <- list()
meta_by_status <- list()
status_levels = c("HC", "IHD")
variable <- "DBP"
for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Diastolic.blood.pressure..mmHg. <- as.numeric(meta$Diastolic.blood.pressure..mmHg.)
  meta <- meta[!is.na(meta$Diastolic.blood.pressure..mmHg.), ]
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, ".eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```


#### Negative Control

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

B <- 1000
set.seed(1)

t_grid_show <- seq(0.01, 0.99, length.out = 100)

meta_df$Diastolic.blood.pressure..mmHg. <- as.numeric(meta_df$Diastolic.blood.pressure..mmHg.)

meta_df <- meta_df %>%
  dplyr::filter(!is.na(Diastolic.blood.pressure..mmHg.))

models <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Diastolic.blood.pressure..mmHg.",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = t_grid_show,
  n_perm = B,
  seed = 123
)

df_tau_coef <- data.frame(
  tau  = models$quantiles,
  coef = models$beta1
)

# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(models$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

df_tau_coef <- cbind(df_tau_coef, perm_mat)

df_perm_long <- df_tau_coef %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

p <- plot_nca(
  result = models,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis",
  xlab = expression(tau),
  ylab = "Coefficient",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "DBP_NegativeControl.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

# Define grid and number of permutations
taus <- seq(0.01, 0.99, length.out = 100)
B <- 1000
set.seed(1)

# Make sure outcome is numeric and remove missing values
meta_df$Diastolic.blood.pressure..mmHg. <- as.numeric(
  meta_df$Diastolic.blood.pressure..mmHg.
)

meta_df <- meta_df %>%
  dplyr::filter(!is.na(Diastolic.blood.pressure..mmHg.))

# Run the new NCA model with confounder adjustment
obs_model <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Diastolic.blood.pressure..mmHg.",
  confounder_cols = c("Depth", "Age", "Gender"),
  basis_df = 6,
  t_grid = taus,
  n_perm = B,
  seed = 123
)

# Observed coefficients
df_tau_coef <- data.frame(
  tau  = obs_model$quantiles,
  coef = obs_model$beta1
)

# Convert permutation matrix to old-style format
# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(obs_model$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

# Bind observed coefficients with permuted coefficients
df_final <- cbind(df_tau_coef, perm_mat)

# Pivot to long format for any downstream use
df_perm_long <- df_final %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# Plot
p <- plot_nca(
  result = obs_model,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis (Blocked Freedman-Lane Permutation)",
  xlab = expression(tau),
  ylab = "Coefficient (DBP)",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p)

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "DBP_NegativeControl_confounder.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```



### Part 10: Wasserstein Spectrum of Left.ventricular
```{r, warning=FALSE}
shannon_by_status <- list()
meta_by_status <- list()
status_levels = c("HC", "IHD")
variable <- "LVEF"
for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Left.ventricular.ejection.fraction.... <- as.numeric(meta$Left.ventricular.ejection.fraction....)
  meta <- meta[!is.na(meta$Left.ventricular.ejection.fraction....), ]
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, ".eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```

#### Negative Control

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

B <- 1000
set.seed(1)

t_grid_show <- seq(0.01, 0.99, length.out = 100)

meta_df$Left.ventricular.ejection.fraction.... <- as.numeric(
  meta_df$Left.ventricular.ejection.fraction....
)

meta_df <- meta_df %>%
  dplyr::filter(!is.na(Left.ventricular.ejection.fraction....))

models <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Left.ventricular.ejection.fraction....",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = t_grid_show,
  n_perm = B,
  seed = 123
)

df_tau_coef <- data.frame(
  tau  = models$quantiles,
  coef = models$beta1
)

# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(models$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

df_tau_coef <- cbind(df_tau_coef, perm_mat)

df_perm_long <- df_tau_coef %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

p <- plot_nca(
  result = models,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis",
  xlab = expression(tau),
  ylab = "Coefficient",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "LeftVentricu_NegativeControl.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```


```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

# Define grid and number of permutations
taus <- seq(0.01, 0.99, length.out = 100)
B <- 1000
set.seed(1)

# Make sure outcome is numeric and remove missing values
meta_df$Left.ventricular.ejection.fraction.... <- as.numeric(
  meta_df$Left.ventricular.ejection.fraction....
)

meta_df <- meta_df %>%
  dplyr::filter(!is.na(Left.ventricular.ejection.fraction....))

# Run the new NCA model with confounder adjustment
obs_model <- wasserstein_spectrum_nca(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Left.ventricular.ejection.fraction....",
  confounder_cols = c("Depth", "Age", "Gender"),
  basis_df = 6,
  t_grid = taus,
  n_perm = B,
  seed = 123
)

# Observed coefficients
df_tau_coef <- data.frame(
  tau  = obs_model$quantiles,
  coef = obs_model$beta1
)

# Convert permutation matrix to old-style format
# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(obs_model$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

# Bind observed coefficients with permuted coefficients
df_final <- cbind(df_tau_coef, perm_mat)

# Pivot to long format for any downstream use
df_perm_long <- df_final %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# Plot
p <- plot_nca(
  result = obs_model,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis (Blocked Freedman-Lane Permutation)",
  xlab = expression(tau),
  ylab = "Coefficient (Left Ventricular Ejection Fraction)",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p)

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "LeftVentricu_NegativeControl_confounder.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```


### Part 11: Wasserstein Spectrum of adiponectin
```{r, warning=FALSE}
shannon_by_status <- list()
meta_by_status <- list()
variable <- "adiponectin"
status_levels = c("MMC", "HC")
for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.adiponectin..mg.L. <- as.numeric(meta$Fasting.plasma.adiponectin..mg.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.adiponectin..mg.L.), ]
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, ".eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})


```

#### Negative Control

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

B <- 1000
set.seed(1)

t_grid_show <- seq(0.01, 0.99, length.out = 100)

# Work on a local copy so meta_df is not modified
df_use <- meta_df %>%
  mutate(
    Fasting.plasma.adiponectin..mg.L. =
      as.numeric(as.character(Fasting.plasma.adiponectin..mg.L.))
  ) %>%
  filter(!is.na(Fasting.plasma.adiponectin..mg.L.))

models <- wasserstein_spectrum_nca(
  df = df_use,
  feature_col = "Shannon",
  outcome_col = "Fasting.plasma.adiponectin..mg.L.",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = t_grid_show,
  n_perm = B,
  seed = 123
)

df_tau_coef <- data.frame(
  tau  = models$quantiles,
  coef = models$beta1
)

# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(models$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

df_tau_coef <- cbind(df_tau_coef, perm_mat)

df_perm_long <- df_tau_coef %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

p <- plot_nca(
  result = models,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis",
  xlab = expression(tau),
  ylab = "Coefficient",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "Adiponectin_NegativeControl.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```


```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

# Define grid and number of permutations
taus <- seq(0.01, 0.99, length.out = 100)
B <- 1000
set.seed(1)

# Work on a local copy so meta_df is not modified
df_use <- meta_df %>%
  mutate(
    Fasting.plasma.adiponectin..mg.L. =
      as.numeric(as.character(Fasting.plasma.adiponectin..mg.L.))
  ) %>%
  filter(!is.na(Fasting.plasma.adiponectin..mg.L.))

# Run the new NCA model with confounder adjustment
obs_model <- wasserstein_spectrum_nca(
  df = df_use,
  feature_col = "Shannon",
  outcome_col = "Fasting.plasma.adiponectin..mg.L.",
  confounder_cols = c("Depth", "Age", "Gender"),
  basis_df = 6,
  t_grid = taus,
  n_perm = B,
  seed = 123
)

# Observed coefficients
df_tau_coef <- data.frame(
  tau  = obs_model$quantiles,
  coef = obs_model$beta1
)

# Convert permutation matrix to old-style format
# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(obs_model$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

# Bind observed coefficients with permuted coefficients
df_final <- cbind(df_tau_coef, perm_mat)

# Pivot to long format for any downstream use
df_perm_long <- df_final %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# Plot
p <- plot_nca(
  result = obs_model,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis (Blocked Freedman-Lane Permutation)",
  xlab = expression(tau),
  ylab = "Coefficient (Adiponectin)",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p)

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "Adiponectin_NegativeControl_confounder.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```

### Part 12: Wasserstein Spectrum of triglycerides
```{r, warning=FALSE}
shannon_by_status <- list()
meta_by_status <- list()
variable <- "triglycerides"
status_levels = c("MMC", "HC")
for (st in status_levels) {
  # Subset phyloseq first
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Compute Shannon for this subset
  shannon_vec <- vegan::diversity(t(otu_table(ps_sub)), index = "shannon")
  
  # Save to meta_df for wasserstein analysis
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- shannon_vec
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.triglycerides..mmol.L. <- as.numeric(meta$Fasting.plasma.triglycerides..mmol.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.triglycerides..mmol.L.), ]
  # Store
  meta_by_status[[st]] <- meta
}


ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, ".eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})


```

#### Negative Control

```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

B <- 1000
set.seed(1)

t_grid_show <- seq(0.01, 0.99, length.out = 100)

# Work on a local copy so meta_df is not modified
df_use <- meta_df %>%
  mutate(
    Fasting.plasma.triglycerides..mmol.L. =
      as.numeric(as.character(Fasting.plasma.triglycerides..mmol.L.))
  ) %>%
  filter(!is.na(Fasting.plasma.triglycerides..mmol.L.))

models <- wasserstein_spectrum_nca(
  df = df_use,
  feature_col = "Shannon",
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = t_grid_show,
  n_perm = B,
  seed = 123
)

df_tau_coef <- data.frame(
  tau  = models$quantiles,
  coef = models$beta1
)

# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(models$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

df_tau_coef <- cbind(df_tau_coef, perm_mat)

df_perm_long <- df_tau_coef %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

p <- plot_nca(
  result = models,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis",
  xlab = expression(tau),
  ylab = "Coefficient",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )

p

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "Triglycerides_NegativeControl.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```


```{r}
# assumes wasserstein_spectrum_nca() and plot_nca() are already defined above

library(tidyverse)

# Define grid and number of permutations
taus <- seq(0.01, 0.99, length.out = 100)
B <- 1000
set.seed(1)

# Work on a local copy so meta_df is not modified
df_use <- meta_df %>%
  mutate(
    Fasting.plasma.triglycerides..mmol.L. =
      as.numeric(as.character(Fasting.plasma.triglycerides..mmol.L.))
  ) %>%
  filter(!is.na(Fasting.plasma.triglycerides..mmol.L.))

# Run the new NCA model with confounder adjustment
obs_model <- wasserstein_spectrum_nca(
  df = df_use,
  feature_col = "Shannon",
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.",
  confounder_cols = c("Depth", "Age", "Gender"),
  basis_df = 6,
  t_grid = taus,
  n_perm = B,
  seed = 123
)

# Observed coefficients
df_tau_coef <- data.frame(
  tau  = obs_model$quantiles,
  coef = obs_model$beta1
)

# Convert permutation matrix to old-style format
# perm_beta1 is B x m, so transpose to m x B
perm_mat <- t(obs_model$perm_beta1)
colnames(perm_mat) <- paste0("Rep_", seq_len(ncol(perm_mat)))

# Bind observed coefficients with permuted coefficients
df_final <- cbind(df_tau_coef, perm_mat)

# Pivot to long format for any downstream use
df_perm_long <- df_final %>%
  select(tau, starts_with("Rep_")) %>%
  pivot_longer(
    cols = starts_with("Rep_"),
    names_to = "Rep",
    values_to = "coef_perm"
  )

# Plot
p <- plot_nca(
  result = obs_model,
  alpha = 0.05,
  n_ticks = 10,
  title = "Negative Control Analysis (Blocked Freedman-Lane Permutation)",
  xlab = expression(tau),
  ylab = "Coefficient (Triglycerides)",
  box_color = "grey80",
  point_color = "red",
  point_shape = 4,
  point_size = 3,
  bound_color = "blue",
  bound_shape = 1,
  bound_size = 1.5
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(p)

dir.create("figures", showWarnings = FALSE, recursive = TRUE)

ggplot2::ggsave(
  filename = file.path("figures", "Triglycerides_NegativeControl_confounder.png"),
  plot = p,
  width = 8, height = 6, units = "in",
  dpi = 300,
  bg = "white"
)
```

## Faith's PD

```{r}
tree = phy_tree(ps_filt)
shannon_vec <- setNames(picante::pd(t(otu_table(ps_filt)), tree, include.root = TRUE)$PD, rownames(t(otu_table(ps_filt))))
```

### Part 1: Three Group Comparison (traditional analysis)

```{r}
meta_df <- data.frame(sample_data(ps_filt))
meta_df$SampleID <- rownames(meta_df)
meta_df$Shannon <- shannon_vec[meta_df$SampleID]
meta_df$Depth <- colSums(otu_table(ps_filt))

kruskal_res <- kruskal.test(Shannon ~ Status, data = meta_df)
p_val <- format.pval(kruskal_res$p.value, digits = 4, eps = .0001)

ggplot(meta_df, aes(x = Status, y = Shannon, fill = Status)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.7) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Alpha Diversity by Group (Shannon Index)",
    y = "Shannon Index",
    x = "Status"
  ) +
  annotate("text", x = 1.5, y = max(meta_df$Shannon) * 1.05,
           label = paste0("Kruskal-Wallis p = ", p_val), size = 4.5) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )

frechet_wasserstein_test(meta_df, 
                         "Shannon", 
                         "Status", 
                         nperm = 1000, 
                         plot = TRUE,
                         seed = 123)

```

### Part 2: Wasserstein Spectrum of HbA1c
```{r}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()
variable <- "HbA1c"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$HbA1c <- as.numeric(meta$HbA1c)
  meta <- meta[!is.na(meta$HbA1c), ]

  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Faith.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```


### Part 3: Wasserstein Spectrum of Status

```{r}
Status.ws <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws$plot

Status.MANOVA <- spectrum_manova(Status.ws)

plot_manova(manova_df = Status.MANOVA,
            type = "both",
            p_cutoff = 0.05,
            line_color = "darkblue",
            ribbon_color = "lightblue",
            title = "Quantile-wise MANOVA",
            base_size = 14)

### adjust sequencing depth

Status.ws2 <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws2$plot

### adjust sequencing depth and other covariates

Status.ws3 <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = c("Depth","Age","HbA1c","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws3$plot


```

### Part 4: Wasserstein Spectrum of BMI
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()
variable <- "BMI"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$BMI..kg.m.. <- as.numeric(meta$BMI..kg.m..)
  meta <- meta[!is.na(meta$BMI..kg.m..), ]
  
  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Faith.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})






```

### Part 5: Wasserstein Spectrum of Waist.circumference
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()

for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$Waist.circumference..cm. <- as.numeric(meta$Waist.circumference..cm.)
  meta <- meta[!is.na(meta$Waist.circumference..cm.), ]
  
  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <- lapply(meta_by_status, function(df_sub) {
  HbA1c.ws <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Waist.circumference..cm.", 
    confounder_cols = NULL,
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

  HbA1c.ws2 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Waist.circumference..cm.", 
    confounder_cols = "Depth",
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.ws3 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Waist.circumference..cm.", 
    confounder_cols = c("Depth", "Age", "Gender"),
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )
})




```

### Part 6: Wasserstein Spectrum of Fasting.plasma.CRP
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()

for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.CRP..mg.L. <- as.numeric(meta$Fasting.plasma.CRP..mg.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.CRP..mg.L.), ]
  
  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <- lapply(meta_by_status, function(df_sub) {
  HbA1c.ws <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Fasting.plasma.CRP..mg.L.", 
    confounder_cols = NULL,
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

  HbA1c.ws2 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Fasting.plasma.CRP..mg.L.", 
    confounder_cols = "Depth",
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.ws3 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Fasting.plasma.CRP..mg.L.", 
    confounder_cols = c("Depth", "Age", "Gender"),
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )
})

```


### Part 7: Wasserstein Spectrum of pro.ANP
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()
variable <- "ANP"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.pro.ANP..pmol.L. <- as.numeric(meta$Fasting.plasma.pro.ANP..pmol.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.pro.ANP..pmol.L.), ]
  
  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Faith.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```

### Part 8: Wasserstein Spectrum of SBP
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()
variable <- "SBP"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$Systolic.blood.pressure..mmHg. <- as.numeric(meta$Systolic.blood.pressure..mmHg.)
  meta <- meta[!is.na(meta$Systolic.blood.pressure..mmHg.), ]
  
  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Faith.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})
```

### Part 9: Wasserstein Spectrum of DBP
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()
variable <- "DBP"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$Diastolic.blood.pressure..mmHg. <- as.numeric(meta$Diastolic.blood.pressure..mmHg.)
  meta <- meta[!is.na(meta$Diastolic.blood.pressure..mmHg.), ]
  
  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Faith.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```

### Part 10: Wasserstein Spectrum of Left.ventricular
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()
variable <- "LVEF"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$Left.ventricular.ejection.fraction.... <- as.numeric(meta$Left.ventricular.ejection.fraction....)
  meta <- meta[!is.na(meta$Left.ventricular.ejection.fraction....), ]
  
  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "Faith.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})
```

### Part 11: Wasserstein Spectrum of adiponectin
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()
variable <- "adiponectin"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.adiponectin..mg.L. <- as.numeric(meta$Fasting.plasma.adiponectin..mg.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.adiponectin..mg.L.), ]
  
  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Faith.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})


```


### Part 12: Wasserstein Spectrum of triglycerides
```{r, warning=FALSE}
status_levels <- unique(sample_data(ps_filt)$Status)
tree <- phy_tree(ps_filt)
meta_by_status <- list()
variable <- "triglycerides"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)

  # Faith's PD (samples as rows → transpose first)
  pd_out <- picante::pd(t(otu_table(ps_sub)), tree, include.root = TRUE)
  pd_vec <- setNames(pd_out$PD, rownames(pd_out))  # Named vector for Faith's PD

  # Get metadata
  meta <- as.data.frame(sample_data(ps_sub))
  meta$Shannon <- pd_vec[rownames(meta)]  # assign PD to Shannon column
  meta$Depth <- colSums(otu_table(ps_sub))  # or rowSums if samples are rows
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.triglycerides..mmol.L. <- as.numeric(meta$Fasting.plasma.triglycerides..mmol.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.triglycerides..mmol.L.), ]
  
  # Store
  meta_by_status[[st]] <- meta
}

# Run wasserstein spectrum models
ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Faith.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})
```

## Chao1

```{r}
shannon_vec <- setNames(apply(t(otu_table(ps_filt)), 1, function(x) vegan::estimateR(x)["S.chao1"]), rownames(t(otu_table(ps_filt))))
```

### Part 1: Three Group Comparison (traditional analysis)

```{r}
meta_df <- data.frame(sample_data(ps_filt))
meta_df$SampleID <- rownames(meta_df)
meta_df$Shannon <- shannon_vec[meta_df$SampleID]
meta_df$Depth <- colSums(otu_table(ps_filt))

kruskal_res <- kruskal.test(Shannon ~ Status, data = meta_df)
p_val <- format.pval(kruskal_res$p.value, digits = 4, eps = .0001)

ggplot(meta_df, aes(x = Status, y = Shannon, fill = Status)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.7) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Alpha Diversity by Group (Shannon Index)",
    y = "Shannon Index",
    x = "Status"
  ) +
  annotate("text", x = 1.5, y = max(meta_df$Shannon) * 1.05,
           label = paste0("Kruskal-Wallis p = ", p_val), size = 4.5) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )

frechet_wasserstein_test(meta_df, 
                         "Shannon", 
                         "Status", 
                         nperm = 1000, 
                         plot = TRUE,
                         seed = 123)

```

### Part 2: Wasserstein Spectrum of HbA1c
```{r}
status_levels <- unique(sample_data(ps_filt)$Status)
meta_by_status <- list()
variable <- "HbA1c"
# Precompute Chao1
chao1_vec <- setNames(
  apply(t(otu_table(ps_filt)), 1, function(x) vegan::estimateR(x)["S.chao1"]),
  rownames(t(otu_table(ps_filt)))
)

for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$HbA1c <- as.numeric(meta$HbA1c)
  meta <- meta[!is.na(meta$HbA1c), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Chao1.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```


### Part 3: Wasserstein Spectrum of Status

```{r}
Status.ws <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws$plot

Status.MANOVA <- spectrum_manova(Status.ws)

plot_manova(manova_df = Status.MANOVA,
            type = "both",
            p_cutoff = 0.05,
            line_color = "darkblue",
            ribbon_color = "lightblue",
            title = "Quantile-wise MANOVA",
            base_size = 14)

### adjust sequencing depth

Status.ws2 <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws2$plot

### adjust sequencing depth and other covariates

Status.ws3 <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = c("Depth","Age","HbA1c","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws3$plot


```

### Part 4: Wasserstein Spectrum of BMI
```{r, warning=FALSE}
variable <- "BMI"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$BMI..kg.m.. <- as.numeric(meta$BMI..kg.m..)
  meta <- meta[!is.na(meta$BMI..kg.m..), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Chao1.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```

### Part 5: Wasserstein Spectrum of Waist.circumference
```{r, warning=FALSE}
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Waist.circumference..cm. <- as.numeric(meta$Waist.circumference..cm.)
  meta <- meta[!is.na(meta$Waist.circumference..cm.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(meta_by_status, function(df_sub) {
  HbA1c.ws <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Waist.circumference..cm.", 
    confounder_cols = NULL,
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

  HbA1c.ws2 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Waist.circumference..cm.", 
    confounder_cols = "Depth",
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.ws3 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Waist.circumference..cm.", 
    confounder_cols = c("Depth", "Age", "Gender"),
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

})
```

### Part 6: Wasserstein Spectrum of Fasting.plasma.CRP
```{r, warning=FALSE}
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.CRP..mg.L. <- as.numeric(meta$Fasting.plasma.CRP..mg.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.CRP..mg.L.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(meta_by_status, function(df_sub) {
  HbA1c.ws <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Fasting.plasma.CRP..mg.L.", 
    confounder_cols = NULL,
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

  HbA1c.ws2 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Fasting.plasma.CRP..mg.L.", 
    confounder_cols = "Depth",
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.ws3 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Fasting.plasma.CRP..mg.L.", 
    confounder_cols = c("Depth", "Age", "Gender"),
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

})

```


### Part 7: Wasserstein Spectrum of pro.ANP
```{r, warning=FALSE}
variable <- "ANP"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.pro.ANP..pmol.L. <- as.numeric(meta$Fasting.plasma.pro.ANP..pmol.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.pro.ANP..pmol.L.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "Chao1.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})



```

### Part 8: Wasserstein Spectrum of SBP
```{r, warning=FALSE}
variable <- "SBP"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Systolic.blood.pressure..mmHg. <- as.numeric(meta$Systolic.blood.pressure..mmHg.)
  meta <- meta[!is.na(meta$Systolic.blood.pressure..mmHg.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Chao1.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})




```

### Part 9: Wasserstein Spectrum of DBP
```{r, warning=FALSE}
variable <- "DBP"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Diastolic.blood.pressure..mmHg. <- as.numeric(meta$Diastolic.blood.pressure..mmHg.)
  meta <- meta[!is.na(meta$Diastolic.blood.pressure..mmHg.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "Chao1.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})


```

### Part 10: Wasserstein Spectrum of Left.ventricular
```{r, warning=FALSE}
variable <- "LVEF"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Left.ventricular.ejection.fraction.... <- as.numeric(meta$Left.ventricular.ejection.fraction....)
  meta <- meta[!is.na(meta$Left.ventricular.ejection.fraction....), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Chao1.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```

### Part 11: Wasserstein Spectrum of adiponectin
```{r, warning=FALSE}
variable <- "adiponectin"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.adiponectin..mg.L. <- as.numeric(meta$Fasting.plasma.adiponectin..mg.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.adiponectin..mg.L.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Chao1.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```


### Part 12: Wasserstein Spectrum of triglycerides
```{r, warning=FALSE}
variable <- "triglycerides"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.triglycerides..mmol.L. <- as.numeric(meta$Fasting.plasma.triglycerides..mmol.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.triglycerides..mmol.L.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_Chao1.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```

## q-PD (q=1)

```{r}
tree = phy_tree(ps_filt)
shannon_vec <-  hill_phylo(comm = t(otu_table(ps_filt)), tree = tree, q = 1, rel_then_pool = FALSE)
```

### Part 1: Three Group Comparison (traditional analysis)

```{r}
meta_df <- data.frame(sample_data(ps_filt))
meta_df$SampleID <- rownames(meta_df)
meta_df$Shannon <- shannon_vec[meta_df$SampleID]
meta_df$Depth <- colSums(otu_table(ps_filt))

kruskal_res <- kruskal.test(Shannon ~ Status, data = meta_df)
p_val <- format.pval(kruskal_res$p.value, digits = 4, eps = .0001)

ggplot(meta_df, aes(x = Status, y = Shannon, fill = Status)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.7) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Alpha Diversity by Group (Shannon Index)",
    y = "Shannon Index",
    x = "Status"
  ) +
  annotate("text", x = 1.5, y = max(meta_df$Shannon) * 1.05,
           label = paste0("Kruskal-Wallis p = ", p_val), size = 4.5) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )

frechet_wasserstein_test(meta_df, 
                         "Shannon", 
                         "Status", 
                         nperm = 1000, 
                         plot = TRUE,
                         seed = 123)

```

### Part 2: Wasserstein Spectrum of HbA1c
```{r}
status_levels <- unique(sample_data(ps_filt)$Status)
meta_by_status <- list()
variable <- "HbA1c"
# Precompute Chao1
chao1_vec <- hill_phylo(comm = t(otu_table(ps_filt)), tree = tree, q = 1, rel_then_pool = FALSE)

for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$HbA1c <- as.numeric(meta$HbA1c)
  meta <- meta[!is.na(meta$HbA1c), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "HbA1c", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_qPD.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})
```


### Part 3: Wasserstein Spectrum of Status

```{r}
Status.ws <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws$plot

Status.MANOVA <- spectrum_manova(Status.ws)

plot_manova(manova_df = Status.MANOVA,
            type = "both",
            p_cutoff = 0.05,
            line_color = "darkblue",
            ribbon_color = "lightblue",
            title = "Quantile-wise MANOVA",
            base_size = 14)

### adjust sequencing depth

Status.ws2 <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws2$plot

### adjust sequencing depth and other covariates

Status.ws3 <- wasserstein_spectrum_multiclass(
  df = meta_df,
  feature_col = "Shannon",
  outcome_col = "Status",
  reference_level = "IHD",
  confounder_cols = c("Depth","Age","HbA1c","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

Status.ws3$plot


```

### Part 4: Wasserstein Spectrum of BMI
```{r, warning=FALSE}
variable <- "BMI"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$BMI..kg.m.. <- as.numeric(meta$BMI..kg.m..)
  meta <- meta[!is.na(meta$BMI..kg.m..), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "BMI..kg.m..", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_qPD.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```

### Part 5: Wasserstein Spectrum of Waist.circumference
```{r, warning=FALSE}
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Waist.circumference..cm. <- as.numeric(meta$Waist.circumference..cm.)
  meta <- meta[!is.na(meta$Waist.circumference..cm.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(meta_by_status, function(df_sub) {
  HbA1c.ws <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Waist.circumference..cm.", 
    confounder_cols = NULL,
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

  HbA1c.ws2 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Waist.circumference..cm.", 
    confounder_cols = "Depth",
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.ws3 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Waist.circumference..cm.", 
    confounder_cols = c("Depth", "Age", "Gender"),
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

})
```

### Part 6: Wasserstein Spectrum of Fasting.plasma.CRP
```{r, warning=FALSE}
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.CRP..mg.L. <- as.numeric(meta$Fasting.plasma.CRP..mg.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.CRP..mg.L.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(meta_by_status, function(df_sub) {
  HbA1c.ws <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Fasting.plasma.CRP..mg.L.", 
    confounder_cols = NULL,
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

  HbA1c.ws2 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Fasting.plasma.CRP..mg.L.", 
    confounder_cols = "Depth",
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

  HbA1c.ws3 <- wasserstein_spectrum(
    df = df_sub, 
    feature_col = "Shannon", 
    outcome_col = "Fasting.plasma.CRP..mg.L.", 
    confounder_cols = c("Depth", "Age", "Gender"),
    basis_df = 6,
    t_grid = seq(0.01, 0.99, length.out = 100),
    alpha = 0.05,
    plot = TRUE,
    seed = 123
  )

})

```


### Part 7: Wasserstein Spectrum of pro.ANP
```{r, warning=FALSE}
variable <- "ANP"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.pro.ANP..pmol.L. <- as.numeric(meta$Fasting.plasma.pro.ANP..pmol.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.pro.ANP..pmol.L.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.pro.ANP..pmol.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_qPD.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})



```

### Part 8: Wasserstein Spectrum of SBP
```{r, warning=FALSE}
variable <- "SBP"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Systolic.blood.pressure..mmHg. <- as.numeric(meta$Systolic.blood.pressure..mmHg.)
  meta <- meta[!is.na(meta$Systolic.blood.pressure..mmHg.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Systolic.blood.pressure..mmHg.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_qPD.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})




```

### Part 9: Wasserstein Spectrum of DBP
```{r, warning=FALSE}
variable <- "DBP"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Diastolic.blood.pressure..mmHg. <- as.numeric(meta$Diastolic.blood.pressure..mmHg.)
  meta <- meta[!is.na(meta$Diastolic.blood.pressure..mmHg.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Diastolic.blood.pressure..mmHg.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_qPD.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})

```

### Part 10: Wasserstein Spectrum of Left.ventricular
```{r, warning=FALSE}
variable <- "LVEF"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Left.ventricular.ejection.fraction.... <- as.numeric(meta$Left.ventricular.ejection.fraction....)
  meta <- meta[!is.na(meta$Left.ventricular.ejection.fraction....), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <- lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Left.ventricular.ejection.fraction....", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_qPD.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})
```

### Part 11: Wasserstein Spectrum of adiponectin
```{r, warning=FALSE}
variable <- "adiponectin"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.adiponectin..mg.L. <- as.numeric(meta$Fasting.plasma.adiponectin..mg.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.adiponectin..mg.L.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.adiponectin..mg.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_qPD.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})
```


### Part 12: Wasserstein Spectrum of triglycerides
```{r, warning=FALSE}
variable <- "triglycerides"
for (st in status_levels) {
  # Subset phyloseq
  ps_sub <- subset_samples(ps_filt, Status == st)
  
  # Sample metadata
  meta <- as.data.frame(sample_data(ps_sub))
  
  # Assign Chao1 richness
  meta$Shannon <- chao1_vec[rownames(meta)]  # storing in 'Shannon' column for modeling
  meta$Depth <- colSums(otu_table(ps_sub))
  meta$SampleID <- rownames(meta)
  meta$Fasting.plasma.triglycerides..mmol.L. <- as.numeric(meta$Fasting.plasma.triglycerides..mmol.L.)
  meta <- meta[!is.na(meta$Fasting.plasma.triglycerides..mmol.L.), ]
  
  meta_by_status[[st]] <- meta
}

# Run Wasserstein spectrum GLS models
ws_list <-  lapply(names(meta_by_status), function(st) {
  df_sub <- meta_by_status[[st]]
  HbA1c.ws <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = NULL,
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

HbA1c.qflt <- quantile_FLT(spectrum_obj = HbA1c.ws, a = 0.2, b = 0.8)

### adjust sequencing depth
HbA1c.ws2 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = "Depth",
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)

### adjust other covariates
HbA1c.ws3 <- wasserstein_spectrum(
  df = df_sub, 
  feature_col = "Shannon", 
  outcome_col = "Fasting.plasma.triglycerides..mmol.L.", 
  confounder_cols = c("Depth","Age","Gender"),
  basis_df = 6,
  t_grid = seq(0.01, 0.99, length.out = 100),
  alpha = 0.05,
  plot = TRUE,
  seed = 123
)
print(HbA1c.ws3$lollipop_plot)
fig_path <- paste0("figures/", st, "_", variable, "_qPD.eps")
ggsave(
  filename = fig_path,
  plot = HbA1c.ws3$lollipop_plot,
  device = cairo_ps,
  width = 7,
  height = 5,
  fallback_resolution = 600
)
})


```









